OpenMPI implementation of Burrows-Wheeler Transform for substring matching applied to genomics

### 2023-08-31 report/report.tex
final fixes

%% The amssymb package provides various useful mathematical symbols
\usepackage{hyperref}
\usepackage{listings}
\usepackage{amssymb}
\usepackage{lipsum}
\usepackage{float}

@@

\section{Performance and scalability analysis}
The parallel implementation of the code has been tested and debugged locally at first in order to make sure everything was working properly. Once obtained a working version having reasonable performance on a local machine, the code has been tested on different types of clusters instanced on Google Cloud Platform in order to evaluate its scalability. \\
The Google Cloud Platform (GCP) offers a suite of cloud computing services including data storage and management, computing, networking, big data and machine learning tools, allowing businesses to build, deploy and scale their applications and services on Google's infrastructure. \\
The performance of the parallel version of the code has been tested and evaluated on the following four clusters in order to investigate the difference in performance on different infrastructures:
\begin{itemize}
   \item \textbf{Fat cluster}: three e2-highmem-8 (8 vCPU, 4 core, 64GB RAM)
   \begin{itemize}
       \item \textbf{Intra-regional}: same region (us-central1-a)
       \item \textbf{Infra-regional}: different regions (us-central1-a, europe-north1-a, asia-east1-b)
   \end{itemize}
   \item \textbf{Light cluster}: six e2-highmem-2 (2 vCPU, 1 core, 16GB RAM)
   \begin{itemize}
       \item \textbf{Intra-regional}: same region (us-central1-a)
       \item \textbf{Infra-regional}: different regions (us-central1-a, us-west2-a, europe-north1-a, europe-west9-a, asia-east1-b, asia-south2-a)
   \end{itemize}
\end{itemize}
The tests have been performed by setting up a hostfile on the master containing the local IP addresses and the number of slots (vCPU) of each node of the cluster and then running the following command \\
\begin{lstlisting}[language=bash]
mpirun --hostfile hostfile -np [N] ./a.out [genome_path] [pattern_path]
\end{lstlisting}
The execution times measured for each execution have been collected in a \emph{.csv} file where a first analysis was performed to compute the speedup and efficiency using excel features. Speedup and efficiency are important metrics needed to assess strong and weak scalability. \\
\textbf{Strong scalability} is the capability of a parallel program to reduce the computational time of a fixed size problem by adding more processors. Strong scalability is important to consider when we deal with a fixed amount of data and we want to speedup the execution of the computation. A simple way to evaluate it is computing the speedup as the ratio between the time t(1) needed to complete a serial task, and the amount of time t(N) needed to complete the same unit of work with N processing elements (parallel task)\cite{hpc scaling}:
$$ Speedup(N) = \frac{t(1)}{t(N)} $$
Whereas, \textbf{weak scalability} is the capability of a parallel program to keep the computational time fixed while increasing the size of the problem. Weak scalability is important to consider when we deal with problems that can increase in size and we want to keep the computation efficient by distributing it on more nodes. A simple way to assess weak scalability is to compute efficiency as the ratio between the time t(1) needed to complete a work unit with 1 processing element, and the amount of time t(N) needed to complete N of the same work units with N processing elements\cite{hpc scaling}:
$$ Efficiency(N) = \frac{t(1)}{t(N)} $$
Efficiency can also be estimated as the ratio between the speedup and the number of processing units used:
$$ Efficiency = \frac{Speedup}{N} $$
In the next subsections the performance on the different types of clusters will be discussed. All the following graphs have been generated from the measures collected using the python script that can be found in the project repository\cite{analysis me}.

@@

\section{Conclusions}
The parallel implementation of the Burrows-Wheeler algorithm for substring matching provided in this project performed well on intra-regional clusters, both fat and light, in terms of both strong and weak scalability, especially for small datasets. Whereas for larger datasets it proved to be less scalable probably due to data transfer overheads. The speedups obtained were in line with what was expected from the a priori analysis, although a bit worse due to communication overheads. \\
On the other hand, performances were not as satisfactory on infra-regional clusters, where the code achieved lower speedup and efficiency both on fat and light clusters compared to the corresponding intra-regional experiments. Moreover, as opposed to the intra-regional scenario, in the infra-regional experiments smaller datasets proved to be less scalable, probably because of the overheads needed to just connect and synchronize machines far from each other.

\begin{thebibliography}{9}
\bibitem{nucleotide} \url{https://en.wikipedia.org/wiki/Nucleotide}
\bibitem{ncbi} \url{https://www.ncbi.nlm.nih.gov/genbank/}
\bibitem{ecoli} \url{https://en.wikipedia.org/wiki/Escherichia_coli}
\bibitem{ecoli dataset} \url{https://www.ncbi.nlm.nih.gov/datasets/taxonomy/562/}
\bibitem{entamoeba} \url{https://en.wikipedia.org/wiki/Entamoeba_invadens}
\bibitem{entamoeba dataset} \url{https://www.ncbi.nlm.nih.gov/datasets/taxonomy/33085/}
\bibitem{formica} \url{https://en.wikipedia.org/wiki/Formica_exsecta}
\bibitem{formica dataset} \url{https://www.ncbi.nlm.nih.gov/datasets/taxonomy/72781/}
\bibitem{geeks}
\url{https://www.geeksforgeeks.org/burrows-wheeler-data-transform-algorithm/}
\bibitem{mreasy} 
\url{https://mr-easy.github.io/2019-12-19-burrows-wheeler-alignment-part-1/}
\bibitem{bwt.h me}
\url{https://github.com/AndreiBlindu/genome-pattern-search-OpenMPI/blob/main/src/utils/bwt.h}
\bibitem{parallel me}
\url{https://github.com/AndreiBlindu/genome-pattern-search-OpenMPI/blob/main/src/parallel/main.cpp}
\bibitem{analysis me}
\url{https://github.com/AndreiBlindu/genome-pattern-search-OpenMPI/blob/main/analysis/analysis.ipynb}
\bibitem{hpc scaling}
\url{https://hpc-wiki.info/hpc/Scaling}
\end{thebibliography}

>>>>

%% The amssymb package provides various useful mathematical symbols
\usepackage[hidelinks]{hyperref}
\usepackage{listings}
\usepackage{amssymb}
\usepackage{lipsum}
\usepackage{float}

@@

\section{Performance and scalability analysis}
The parallel implementation of the code has been tested and debugged locally at first in order to make sure everything was working properly. Once obtained a working version having reasonable performance on a local machine, the code has been tested on different types of clusters instanced on Google Cloud Platform in order to evaluate its scalability. \\
The Google Cloud Platform (GCP) offers a suite of cloud computing services including data storage and management, computing, networking, big data and machine learning tools, allowing businesses to build, deploy and scale their applications and services on Google's infrastructure. \\
The performance of the parallel version of the code has been tested and evaluated on the following four clusters in order to investigate the difference in performance on different infrastructures:
\begin{itemize}
   \item \textbf{Fat cluster}: three e2-highmem-8 (8 vCPU, 4 core, 64GB RAM)
   \begin{itemize}
       \item \textbf{Intra-regional}: same region (us-central1-a)
       \item \textbf{Infra-regional}: different regions (us-central1-a, europe-north1-a, asia-east1-b)
   \end{itemize}
   \item \textbf{Light cluster}: six e2-highmem-2 (2 vCPU, 1 core, 16GB RAM)
   \begin{itemize}
       \item \textbf{Intra-regional}: same region (us-central1-a)
       \item \textbf{Infra-regional}: different regions (us-central1-a, us-west2-a, europe-north1-a, europe-west9-a, asia-east1-b, asia-south2-a)
   \end{itemize}
\end{itemize}
The tests have been performed by setting up a hostfile on the master containing the local IP addresses and the number of slots (vCPU) of each node of the cluster and then running the following command \\
\begin{lstlisting}[language=bash]
mpirun --hostfile hostfile -np [N] ./a.out [genome_path] [pattern_path]
\end{lstlisting}
The execution times measured for each execution have been collected in a \emph{.csv} file where a first analysis was performed to compute the speedup and efficiency using excel features (the files containing the results of the experiments can be found in the project repository under the \emph{tests} directory\cite{tests me}). Speedup and efficiency are important metrics needed to assess strong and weak scalability. \\
\textbf{Strong scalability} is the capability of a parallel program to reduce the computational time of a fixed size problem by adding more processors. Strong scalability is important to consider when we deal with a fixed amount of data and we want to speedup the execution of the computation. A simple way to evaluate it is computing the speedup as the ratio between the time t(1) needed to complete a serial task, and the amount of time t(N) needed to complete the same unit of work with N processing elements (parallel task)\cite{hpc scaling}:
$$ Speedup(N) = \frac{t(1)}{t(N)} $$
Whereas, \textbf{weak scalability} is the capability of a parallel program to keep the computational time fixed while increasing the size of the problem. Weak scalability is important to consider when we deal with problems that can increase in size and we want to keep the computation efficient by distributing it on more nodes. A simple way to assess weak scalability is to compute efficiency as the ratio between the time t(1) needed to complete a work unit with 1 processing element, and the amount of time t(N) needed to complete N of the same work units with N processing elements\cite{hpc scaling}:
$$ Efficiency(N) = \frac{t(1)}{t(N)} $$
Efficiency can also be estimated as the ratio between the speedup and the number of processing units used:
$$ Efficiency = \frac{Speedup}{N} $$
In the next subsections the performance on the different types of clusters will be discussed. All the following graphs have been generated from the measures collected using the python script that can be found in the project repository\cite{analysis me}.

@@

\section{Conclusions}
The parallel implementation of the Burrows-Wheeler algorithm for substring matching provided in this project performed well on intra-regional clusters, both fat and light, in terms of both strong and weak scalability, especially for small datasets. Whereas for larger datasets it proved to be less scalable probably due to data transfer overheads. The speedups obtained were in line with what was expected from the a priori analysis, although a bit worse due to communication overheads. \\
On the other hand, performances were not as satisfactory on infra-regional clusters, where the code achieved lower speedup and efficiency both on fat and light clusters compared to the corresponding intra-regional experiments. Moreover, as opposed to the intra-regional scenario, in the infra-regional experiments smaller datasets proved to be less scalable, probably because of the overheads needed to just connect and synchronize machines far from each other. \\
It could be interesting trying to find a way to efficiently parallelize also the reading and prepreocessing of the datasets in order to investigate the possible improvements in performance with respect to this implementation in the different conditions discussed.

\def\UrlBreaks{\do\/\do-}
\begin{thebibliography}{9}
\bibitem{nucleotide} \url{https://en.wikipedia.org/wiki/Nucleotide}
\bibitem{ncbi} \url{https://www.ncbi.nlm.nih.gov/genbank/}
\bibitem{ecoli} \url{https://en.wikipedia.org/wiki/Escherichia_coli}
\bibitem{ecoli dataset} \url{https://www.ncbi.nlm.nih.gov/datasets/taxonomy/562/}
\bibitem{entamoeba} \url{https://en.wikipedia.org/wiki/Entamoeba_invadens}
\bibitem{entamoeba dataset} \url{https://www.ncbi.nlm.nih.gov/datasets/taxonomy/33085/}
\bibitem{formica} \url{https://en.wikipedia.org/wiki/Formica_exsecta}
\bibitem{formica dataset} \url{https://www.ncbi.nlm.nih.gov/datasets/taxonomy/72781/}
\bibitem{geeks}
\url{https://www.geeksforgeeks.org/burrows-wheeler-data-transform-algorithm/}
\bibitem{mreasy} 
\url{https://mr-easy.github.io/2019-12-19-burrows-wheeler-alignment-part-1/}
\bibitem{bwt.h me}
\url{https://github.com/AndreiBlindu/genome-pattern-search-OpenMPI/blob/main/src/utils/bwt.h}
\bibitem{parallel me}
\url{https://github.com/AndreiBlindu/genome-pattern-search-OpenMPI/blob/main/src/parallel/main.cpp}
\bibitem{tests me}
\url{https://github.com/AndreiBlindu/genome-pattern-search-OpenMPI/tree/main/src/parallel/GCP/tests}
\bibitem{analysis me}
\url{https://github.com/AndreiBlindu/genome-pattern-search-OpenMPI/blob/main/analysis/analysis.ipynb}
\bibitem{hpc scaling}
\url{https://hpc-wiki.info/hpc/Scaling}
\end{thebibliography}
